---
title: "Two Tracks, One Athlete: Return-to-School vs Return-to-Sport Staging"
description: "Consensus guidelines run return-to-learn and return-to-sport as parallel tracks—but use conflicting abbreviations, different clocks, and asymmetric real-world implementation. This piece maps both checklists and where aerobic prescription must follow the school leg, not sport clearance alone."
date: "2026-06-19"
series: "Foundation Stack"
installment: pillar2-004
pillar: methods
audience: [clinician, researcher, parent]
categories: [foundation-stack, PPCS, methods, rtl-rtp, staging]
---


::: {.callout-caution}
**Not medical advice. Educational use only. Data from published literature.**
:::

Pulling return-to-learn and return-to-sport checklists side by side, I expected one timeline. I found two vocabularies—**RTS means school in one Canadian protocol and sport in Berlin**—two median clocks (35 vs 38 days in one youth cohort), and chart data saying we still release **both** tracks too early, often on different schedules [PMID: 31396150](https://pubmed.ncbi.nlm.nih.gov/31396150/) [PMID: 24925965](https://pubmed.ncbi.nlm.nih.gov/24925965/).

That is the problem this installment maps.

**What this installment is—and is not.** This is a **methods** piece in the Foundation Stack series. It does **not** replace CISG appendices or your district’s return-to-play law. It maps **two parallel staging checklists** (RTL and RTP), shows where they attach to [pillar2-002](../pillar2-002-no-bctt-prescription-paths/index.html) intensity and [pillar2-003](../pillar2-003-twenty-minutes-from-where/index.html) volume, and flags **when published recovery numbers cannot be compared**. Athletic trainers and school staff will find the most actionable material in [Section 4](#resource-limited-settings) and the [dual-track SOP](#sop-checklist-dual-track). Researchers and trainees will find the endpoint and chart-review critique in [Section 3](#how-to-read-the-recovery-numbers-without-merging-the-tracks)—skippable on first read if you only need staging labels for tomorrow’s clinic.

The [second installment](../pillar2-002-no-bctt-prescription-paths/index.html) asked how to set a sub-symptom heart-rate band when there is no treadmill. The [third](../pillar2-003-twenty-minutes-from-where/index.html) asked how many minutes and sessions fit a school week. Both assumed the **staging question** was mostly settled: once symptoms allow, you advance. In clinic, the adolescent often carries **two** advance rules at once—what the school will permit and what the sport program will permit—and those rules are not written in the same document.

#### 1. Two checklists, one student

Sport-related concussion management now routinely describes **graduated return-to-learn** (RTL) and **graduated return-to-sport** (RTP) as parallel processes. Berlin 2017 was the first International Conference on Concussion in Sport (CISG) consensus to publish paired strategies: a **four-step** return-to-school progression and a **six-step** return-to-sport progression in the same statement [PMID: 28446457](https://pubmed.ncbi.nlm.nih.gov/28446457/). Amsterdam 2022 retained that skeleton, refined symptom rules (mild and brief exacerbation on a 0–10 scale), and stated explicitly that RTL and RTP **may occur in parallel** while **full RTL should precede unrestricted RTP** [PMID: 37316210](https://pubmed.ncbi.nlm.nih.gov/37316210/).

The logic sounds coordinated. The operational reality often is not.

A common failure mode: the athlete is cleared for **light aerobic work or non-contact drills** on the sport track while still on **partial school attendance or formal accommodations** on the learning track—or the reverse, full days at school while still symptom-limited for sport. Neither is automatically wrong; guidelines allow overlap in early stages. What breaks down is **documentation and communication**: the school sees one stage label, the trainer another, and the family hears “cleared to exercise” without hearing “not cleared for contact” or “still on reduced screen time.”

Canadian chart reviews quantify the asymmetry. Carson et al. found **43.5%** of concussion cases returned to sport too soon and **44.7%** returned to school too soon in a 2006–2011 cohort [PMID: 24925965](https://pubmed.ncbi.nlm.nih.gov/24925965/). A follow-up review (2011–2016) showed premature RTP relapse falling (**26.7%**) while premature RTL relapse remained high (**42.6%**)—consistent with structured RTP reaching clinics before structured RTL [PMID: 35292472](https://pubmed.ncbi.nlm.nih.gov/35292472/). Meta-analytic work in athletes reports mean days to RTL **8.3** versus RTS **19.8**, with **93%** achieving full RTL by 10 days without new academic support—yet cohort medians for **full** return in youth can run **35 days (school)** and **38 days (sport)** when stage-by-stage progression is tracked [PMID: 37316183](https://pubmed.ncbi.nlm.nih.gov/37316183/) [PMID: 31396150](https://pubmed.ncbi.nlm.nih.gov/31396150/). Those numbers measure different endpoints; comparing them without a shared definition is its own error mode.

For aerobic prescription specifically, the mismatch matters. Pillar2-002 and pillar2-003 discuss **bands and minutes** on the assumption that cognitive load is negotiable. If the student is still in RTL step 1–3—homework bursts, partial days, screen limits—adding even “sub-symptom” walking or cycling on the sport track can **stack** physical and cognitive load in the same 24 hours. The limiting track may be school, not the heart-rate formula.

##### Why the school column is not “soft” sport advice

The sport track dominates conference tables and athletic-trainer workflows. The school track has its own literature—not a footnote to RTP.

Halstead et al.’s AAP guidance on **return to learning** describes post-concussion classroom difficulty as common and calls for medical-home coordination with schools on environmental and academic adjustments [PMID: 24163302](https://pubmed.ncbi.nlm.nih.gov/24163302/). Purcell, Davis, and Gioia’s systematic review on **return to school** after sport-related concussion reports that **17–73%** of students received accommodations or had difficulty with school return—wide heterogeneity, but not a rare edge case [PMID: 29500251](https://pubmed.ncbi.nlm.nih.gov/29500251/). Factors in that review include symptom burden, rest timing, age/grade, and **course load**—variables absent from most aerobic-exercise trial arms.

Canadian primary-care writing on **returning to school** emphasizes staged re-entry and communication between clinic and classroom—not assuming full attendance on calendar alone [PMID: 37315976](https://pubmed.ncbi.nlm.nih.gov/37315976/). REAP-style workflows explicitly bridge remove–educate–adjust–pace communication between emergency, medical, and school stakeholders [PMID: 26589561](https://pubmed.ncbi.nlm.nih.gov/26589561/). When pillar2-004 asks where aerobic prescription “attaches,” the honest answer is: **often on the sport leg in clinic notes**, while the **evidence for load management** is split across both legs.

DeMatteo’s systematic review adds a sobering implementation note: RTA/RTS protocols are widely adopted, yet **paediatric efficacy evidence remains thin**—supporting 1–2 days of rest then graduated progression, but not proving that any one staging checklist is validated in trial [PMID: 32153982](https://pubmed.ncbi.nlm.nih.gov/32153982/). Staging is therefore as much a **documentation and coordination** problem as a physiology problem.

::: {.callout-note}
## Abbreviation lock (read before the table)

Different research groups label the **same idea** differently—not because the field is unusable, but because school and sport tracks were formalized by separate literature streams that merged only recently (Berlin 2017 onward).

| Framework | School / learn track | Sport / activity track |
|-----------|----------------------|-------------------------|
| **CISG Berlin & Amsterdam** | RTL (return to learn) | RTP / RTS (return to sport) |
| **DeMatteo / CanChild protocols** | **RTS** = return to **school** | **RTA** = return to **activity** (play/sport) |
| **Canadian chart reviews** | RTL | RTP |

**This article uses RTL + RTP in prose.** When citing DeMatteo, translate once: *RTS (DeMatteo) = school*. One shared row in your note prevents the collision from reaching families.
:::

#### 2. Four-step RTL vs six-step RTP (CISG)

*Table note:* Step labels and activities below follow **Amsterdam 2022** appendices (current CISG) [PMID: 37316210](https://pubmed.ncbi.nlm.nih.gov/37316210/). Berlin 2017 used the same **four-step RTL / six-step RTP** structure with closely related activity descriptions [PMID: 28446457](https://pubmed.ncbi.nlm.nih.gov/28446457/). Both recommend **24–48 hours** of relative rest before step 1; Amsterdam allows symptom-limited activity within **24 hours** of injury with defined exacerbation limits. Minimum **24 hours per step** when advancing without symptom worsening.

##### Return-to-learn (RTL) — four steps

| Step | Mental activity (aim) | Activity at each step | Goal |
|------|------------------------|------------------------|------|
| **1** | Daily activities without more than **mild, brief** symptom exacerbation\* | Typical day activities (e.g., reading) with **minimized screen time**; start **5–15 min** and increase gradually | Gradual return to typical activities |
| **2** | School activities | Homework, reading, or other cognitive work **outside** the classroom | Increase tolerance to cognitive work |
| **3** | Return to school **part time** | Partial school day and/or **more rest breaks** during the day | Increase academic participation |
| **4** | Return to school **full time** | Progress until a full day is tolerated without more than mild exacerbation\* | Full academic load; catch up on missed work |

\* **Mild and brief exacerbation (Amsterdam):** increase of **≤2 points** on a 0–10 symptom scale for **<1 hour** compared with pre-activity baseline [PMID: 37316210](https://pubmed.ncbi.nlm.nih.gov/37316210/).

##### Return-to-sport (RTP) — six steps

| Step | Exercise strategy | Activity | Goal |
|------|-------------------|----------|------|
| **1** | Symptom-limited activity | Daily activities that do not exacerbate symptoms (e.g., walking) | Gradual reintroduction of work/school |
| **2** | Aerobic exercise | **2A—Light** (~≤55% age-predicted max HR) then **2B—Moderate** (~≤70% max HR): walking or stationary cycling; may add **light** resistance if symptoms stay mild\* | Increase heart rate |
| **3** | Individual sport-specific exercise | Drills **away from team**; **no head-impact risk** (medical clearance if inadvertent head impact possible) | Movement, change of direction |
| **4** | Non-contact training drills | High-intensity drills (e.g., passing); may rejoin **team environment** | Usual exercise intensity, coordination, thinking load |
| **5** | Full-contact practice | Normal training including contact | Confidence; coaching assessment |
| **6** | Return to sport | Normal game play | Unrestricted competition |

**Amsterdam gate for steps 4–6:** begin only after **resolution of symptoms**, cognitive abnormalities, and other concussion-related findings **including with exertion** [PMID: 37316210](https://pubmed.ncbi.nlm.nih.gov/37316210/). **Berlin gate for children:** do **not** return to sport until **successfully returned to school**, while **early symptom-limited physical activity** remains appropriate [PMID: 28446457](https://pubmed.ncbi.nlm.nih.gov/28446457/). Davis et al.'s child-management review similarly recommends schools **not** return to sport until successful return to school, with early symptom-limited activity [PMID: 28455361](https://pubmed.ncbi.nlm.nih.gov/28455361/).

##### Where the tracks meet (coordination rules)

| Source | Parallel tracks? | School-before-contact rule | Evidence strength |
|--------|------------------|----------------------------|-------------------|
| **Amsterdam 2022** | RTL and RTP **may run in parallel** | **Complete full RTL** before **unrestricted** RTP; youth should reach **full-time school** before RTP **steps 4–6** | Consensus + systematic review [37316210](https://pubmed.ncbi.nlm.nih.gov/37316210/) |
| **Berlin 2017** | Early physical activity while symptomatic | **No sport** until **successful return to school** | Consensus [28446457](https://pubmed.ncbi.nlm.nih.gov/28446457/) |
| **DeMatteo 2019 cohort** | **Stages 1–3** of school (RTS) and activity (RTA) **simultaneously** | **RTA stage 6** (full contact) delayed until **full, successful school reintegration** (RTS) | Prospective cohort *n*=139 [31396150](https://pubmed.ncbi.nlm.nih.gov/31396150/) |
| **DeMatteo 2020 SR** | Protocols adopted in practice | **1–2 days** rest then graduated RTA/**RTS** (DeMatteo labels); **little paediatric efficacy evidence** for either protocol | Systematic review [32153982](https://pubmed.ncbi.nlm.nih.gov/32153982/) |

**Reading the overlap:** Amsterdam RTP **step 1** explicitly reintroduces daily activity—including school-related load—while RTL **steps 1–2** still cap screen time and homework tolerance. Sport **step 2** is where published **%HRmax bands** (pillar2-002) attach: light then moderate aerobic work. That can be **allowed on the sport checklist** while RTL **step 3** (partial school) is still active—precisely the zone where families need both labels on one plan.

#### 3. How to read the recovery numbers (without merging the tracks) {#how-to-read-the-recovery-numbers-without-merging-the-tracks}

Section 2 lists steps; the literature quantifying those steps uses **different rulers**. Three statistical habits—none of them exotic—explain most double-counting errors when RTL and RTP data sit in the same conversation.

##### Three numbers that do not compare

| Number (source) | What it actually measures | Common misread |
|-----------------|---------------------------|----------------|
| **Mean ~8.3 d to RTL** (athletes; meta-analysis) [37316183](https://pubmed.ncbi.nlm.nih.gov/37316183/) | Average **days until RTL** across heterogeneous studies | “Most kids are back to full school in eight days” |
| **93% full RTL by 10 d** (same meta-analysis) [37316183](https://pubmed.ncbi.nlm.nih.gov/37316183/) | Proportion achieving **full RTL without new academic support** by day 10 | Same as row above—but endpoint is **full**, not first step |
| **Median 35 d (school) / 38 d (sport)** (youth cohort; stage tracking) [31396150](https://pubmed.ncbi.nlm.nih.gov/31396150/) | **Full** return on DeMatteo school (RTS) and activity (RTA) protocols | “Meta says 8 days; cohort says 35—someone is wrong” |
| **43.5% / 44.7% returned too soon** (chart review) [24925965](https://pubmed.ncbi.nlm.nih.gov/24925965/) | **Marginal** proportion of all cases judged premature on RTP / RTL | “If we advance sport, 43% relapse” (not supported without a 2×2 table) |
| **26.7% / 42.6% premature with relapse** (later chart review) [35292472](https://pubmed.ncbi.nlm.nih.gov/35292472/) | **Marginal** counts of cases where premature RTP / RTL **and** relapse co-occur in chart documentation | Confused with **P(relapse \| premature)** |

**Takeaway:** Means from meta-analyses, medians from single cohorts, and percentages from chart reviews answer **different questions** on **different endpoints**. Collapsing them into one “recovery timeline” is a methods error—not a disagreement among experts.

##### Figure 1 — Recovery clocks measure different endpoints {#fig-recovery-clocks}

```{r}
#| label: fig-recovery-clocks
#| fig-width: 8.5
#| fig-height: 5
#| echo: false
#| warning: false
#| message: false

library(ggplot2)

fs_source_theme <- function() {
 candidates <- c(
 file.path(dirname(knitr::current_input(dir = TRUE)), "..", "_shared", "fs-ggplot-theme.R"),
 path.expand("~/Documents/guanglab-foundation-stack/posts/_shared/fs-ggplot-theme.R")
 )
 path <- candidates[file.exists(candidates)][1]
 if (!is.na(path)) source(path, local = parent.frame())
}
fs_source_theme()

clock_df <- data.frame(
 label = c(
 "Mean days to RTL",
 "Mean days to RTS",
 "Median full return to school",
 "Median full return to sport"
 ),
 days = c(8.3, 19.8, 35, 38),
 track = c("RTL (school)", "RTP (sport)", "RTL (school)", "RTP (sport)"),
 family = c("Meta-analysis mean", "Meta-analysis mean", "Cohort median", "Cohort median"),
 stringsAsFactors = FALSE
)

clock_df$label <- factor(clock_df$label, levels = rev(clock_df$label))

ggplot(clock_df, aes(x = days, y = label, fill = track)) +
 geom_col(width = 0.62, alpha = 0.92, color = "grey25", linewidth = 0.25) +
 geom_text(aes(label = days), hjust = -0.1, size = 3.6, fontface = "bold", color = FS_COL_INK) +
 annotate(
 "text", x = 40, y = 4.35, label = "Do not connect into one patient timeline",
 size = 3.2, fontface = "italic", color = FS_COL_ALERT
 ) +
 scale_fill_manual(
 values = c("RTL (school)" = FS_COL_RTL, "RTP (sport)" = FS_COL_RTP),
 name = NULL
 ) +
 scale_x_continuous(limits = c(0, 42), expand = expansion(mult = c(0, 0.12))) +
 labs(
 title = "Published recovery clocks use different endpoints",
 subtitle = "Meta-analytic means (athletes) vs cohort medians (full protocol completion, youth)",
 x = "Days reported in source paper",
 y = NULL,
 caption = "Putukian et al. 2023 meta-analysis [PMID: 37316183]; DeMatteo et al. 2019 cohort [PMID: 31396150].\nBars are not standardized to one definition of “return.”"
 ) +
 theme_fs(base_size = 11)
```

**Figure 1.** Bars visualize **four published summary statistics**, not one recovery curve. Meta-analytic **means** describe aggregated study endpoints in athletes; cohort **medians** describe **full** protocol completion in one youth sample. Section 3 argues against quoting any single bar as “typical recovery.”

##### Figure 2 — Sport vs school track in chart reviews (different metrics) {#fig-carson-tracks}

```{r}
#| label: fig-carson-tracks
#| fig-width: 8.5
#| fig-height: 5
#| echo: false
#| warning: false
#| message: false

library(ggplot2)
fs_source_theme()

carson_df <- data.frame(
 metric = rep(c("Premature return\n(too soon)", "Relapse after\npremature return"), each = 2),
 track = rep(c("RTP (sport)", "RTL (school)"), 2),
 pct = c(43.5, 44.7, 26.7, 42.6),
 era = rep(c("Carson 2014", "Carson 2022"), each = 2),
 stringsAsFactors = FALSE
)

carson_df$track <- factor(carson_df$track, levels = c("RTP (sport)", "RTL (school)"))
carson_df$era <- factor(carson_df$era, levels = c("Carson 2014", "Carson 2022"))

ggplot(carson_df, aes(x = track, y = pct, group = metric, fill = track, alpha = metric)) +
 geom_col(position = position_dodge(width = 0.72), width = 0.64, color = "grey25", linewidth = 0.25) +
 geom_text(
 aes(label = paste0(pct, "%")),
 position = position_dodge(width = 0.72),
 vjust = -0.45,
 size = 3.5,
 fontface = "bold",
 color = FS_COL_INK
 ) +
 facet_wrap(~ era, ncol = 2) +
 scale_fill_manual(
 values = c("RTP (sport)" = FS_COL_RTP, "RTL (school)" = FS_COL_RTL),
 guide = "none"
 ) +
 scale_alpha_manual(
 values = c("Premature return\n(too soon)" = 0.55, "Relapse after\npremature return" = 1),
 name = NULL
 ) +
 scale_y_continuous(limits = c(0, 52), expand = expansion(mult = c(0, 0.06))) +
 labs(
 title = "Canadian chart reviews: asymmetric sport vs school implementation",
 subtitle = "2014 = premature return rates; 2022 = chart-documented relapse after premature return (different metrics)",
 x = NULL,
 y = "Percent of concussion cases in chart review",
 caption = "Carson et al. 2014 [PMID: 24925965]; Carson et al. 2022 [PMID: 35292472].\nCompare tracks within each panel—not across panels as identical outcomes."
 ) +
 theme_fs(base_size = 11)
```

**Figure 2.** Carson’s two chart reviews use **different outcome labels** across eras. The pattern worth retaining: **school-track problems remain visible even when sport-track premature relapse improves**—consistent with RTL tooling lagging RTP in many offices [PMID: 35292472](https://pubmed.ncbi.nlm.nih.gov/35292472/).

##### Marginal vs conditional proportions

Carson-style chart reviews are essential for showing **implementation gaps**, but their percentages are usually **of all concussion cases**, not **of athletes who were cleared early**.

Without published cell counts (premature × relapse), you cannot invert a marginal percentage into a personal risk statement. A defensible clinical read is associative: **premature advance on either track still appears common**, and **relapse after premature advance still appears in charts** [24925965](https://pubmed.ncbi.nlm.nih.gov/24925965/) [35292472](https://pubmed.ncbi.nlm.nih.gov/35292472/)—not that a specific RTL step confers a known relapse probability.

##### Mean, median, and heavy tails

Recovery days are typically **right-skewed**: most adolescents recover within weeks; a minority extend much longer. **Means** (meta-analysis) and **medians** (DeMatteo cohort) can both be correct yet far apart if long tails pull means or if cohorts define “return” differently.

Putukian et al. also report **I² ≈ 99%** for days-to-RTL and days-to-RTS outcomes [37316183](https://pubmed.ncbi.nlm.nih.gov/37316183/)—meaning study definitions and populations differ enormously. Figure 1 shows why a single “days to return” number misleads: wide confidence intervals on a meta-analytic **mean** describe uncertainty about the **average across studies**, not the spread for **the next patient**. Individual clocks will often sit outside any single quoted band.

##### Dual-track state is a cross-classification—not a single stage label

An athlete at **RTL step 3** (partial school) and **RTP step 2B** (moderate aerobic) occupies one **cell** in a 4×6 grid, not “step 3 overall.” Guidelines permit some combinations; Amsterdam **restricts RTP steps 4–6** until symptom resolution with exertion and expects **full-time school** before those high-risk sport stages [37316210](https://pubmed.ncbi.nlm.nih.gov/37316210/).

| RTL (school) | RTP (sport) zone | Guideline read |
|--------------|------------------|----------------|
| 1–2 (home / homework) | 1 (symptom-limited daily activity) | Overlap expected; cap **total** cognitive + physical load same day |
| 3 (part-time school) | 2A–2B (light–moderate aerobic) | **Common clinic zone**—sport may be “allowed” while school still formalized |
| 4 (full-time school) | 3 (individual sport-specific, no head impact) | Aerobic prescription (pillar2-002) often active; contact still off |
| 4 (full-time, no accommodations) | 4–6 (team drills → contact → play) | Amsterdam: high-risk sport after **full RTL** + exertional symptom clearance |

Document **both coordinates**. “Cleared to exercise” without an RTL label is an incomplete state vector.

##### Three operational disagreements (expanded)

**Disagreement 1 — Sequence:** Consensus text emphasizes **school success before unrestricted sport** [28446457](https://pubmed.ncbi.nlm.nih.gov/28446457/) [37316210](https://pubmed.ncbi.nlm.nih.gov/37316210/). DeMatteo’s cohort shows **simultaneous early-stage progression** is common and can be safe for non-contact phases [31396150](https://pubmed.ncbi.nlm.nih.gov/31396150/). Chart data suggest **RTP structured before RTL** in many offices [35292472](https://pubmed.ncbi.nlm.nih.gov/35292472/). The disagreement is not “school first vs parallel ever”—it is **where the hard gate sits** (contact sport vs any activity) and **who documents the school leg**.

**Disagreement 2 — Clocks:** Meta means (~8 d RTL, ~20 d RTS) describe **aggregated study endpoints** in athletes [37316183](https://pubmed.ncbi.nlm.nih.gov/37316183/). Youth medians (~35 / ~38 d) describe **full protocol completion** in one prospective sample [31396150](https://pubmed.ncbi.nlm.nih.gov/31396150/). Quoting the fast number to reassure families while the school still runs a six-week accommodation plan misaligns expectations.

**Disagreement 3 — Prescription attachment:** Sub-symptom **%HRmax** and **weekly minutes** (pillar2-002/003) attach naturally to **RTP step 2**. **Screen limits, exam deferrals, and partial days** attach to **RTL**. If only the sport prescription is updated at follow-up, the family may **stack** cognitive and aerobic load on the same afternoon—symptoms spike, and both tracks blame the wrong lever.

**What would settle disagreement 1?** Pragmatic trials comparing coordinated dual-track documentation vs single-track RTP-focused follow-up, with **same-day** symptom and adherence reporting on **both** legs—not only sport clearance.

**What would settle disagreement 2?** Standardized reporting in cohort studies: **date of RTL step 1–4** and **RTP step 1–6** in the same row, with **symptom scores after combined daily load**—not separate sport-only exertion tests.

::: {.callout-caution}
**Not medical advice. Educational use only. Data from published literature.**
:::

#### 4. If your setting looks like this {#resource-limited-settings}

pillar2-002 asked how to set **intensity** without BCTT. pillar2-003 asked how to log **weekly minutes**. This installment asks what happens when the **school checklist** and **sport checklist** are updated by different adults on different days.

##### Setting matrix (dual-track)

::: {=html}
<div class="fs-scenario-grid">
<div class="fs-scenario-card">
<span class="fs-tag">Setting A</span>
<h4>School athletic trainer + classroom</h4>
<ul>
<li>AT runs RTP steps; counselor/nurse holds RTL accommodations</li>
<li>Clearance notes often sport-weighted</li>
<li>School may not see HR band or walk prescription</li>
</ul>
<p class="fs-path">Risk: RTP step 2 logged; RTL step 3 invisible on same plan</p>
</div>
<div class="fs-scenario-card">
<span class="fs-tag">Setting B</span>
<h4>Sports medicine clinic</h4>
<ul>
<li>Visit note documents RTP; school forms lag</li>
<li>May prescribe aerobic dose without calling school</li>
<li>REAP / Living Guideline tools exist but are separate PDFs [26589561](https://pubmed.ncbi.nlm.nih.gov/26589561/) [34777657](https://pubmed.ncbi.nlm.nih.gov/34777657/)</li>
</ul>
<p class="fs-path">Default: export <strong>both</strong> step labels to school and coach same visit</p>
</div>
<div class="fs-scenario-card">
<span class="fs-tag">Setting C</span>
<h4>Primary care + parent supervision</h4>
<ul>
<li>Parent hears “walk 20 minutes” from clinic</li>
<li>School emails accommodation list separately</li>
<li>No shared staging document</li>
</ul>
<p class="fs-path">Default: one home log with RTL + RTP columns</p>
</div>
</div>
:::

##### Vignette

A **15-year-old JV guard** took an elbow to the head under the basket **18 days** into the winter season—still in the window where most adolescents recover, but already juggling **missed labs**, a team conditioning expectation, and midterm prep. After urgent-care clearance, three documents diverged:

1. **Sports medicine follow-up (week 2)** — discharge sheet: symptom-limited walk **15–20 min**, **5 days/week**, stop if symptoms rise **≥2/10** from walk start (pillar2-002 symptom-cap path; no BCTT). Parent interprets this as permission to “get back in shape.”

2. **School accommodation letter (week 2)** — counselor lists **RTL step 3** items: **full attendance** with **rest breaks**, **tests deferred one week**, **no loud assemblies**, **PE excused**, preferential seating away from overhead lights. Not a half-day schedule—many districts keep students in building while limiting cognitive load.

3. **Athletic trainer team form (week 3)** — checkbox **“RTP stage 2: light aerobic exercise.”** Coach reads this as green light for tomorrow’s pre-practice jog with the team. The form has **no RTL column** and **no line for same-day homework load**.

**Wednesday:** She completes a **solo 18-minute walk** after school (symptoms **2/10 → 3/10**, acceptable). That evening she makes up a **biology lab write-up** (~50 minutes, laptop, bright screen)—symptoms **3/10 → 6/10** until bedtime. Thursday AM fog is worse; she sits out conditioning. Coach says she “failed RTP”; teacher notes she “was fine in class until she overdid exercise.”

The failure mode is **siloed narratives**, not a stopwatch error. Amsterdam’s mild-exacerbation rule applies to **each activity class**—but families experience **one cumulative day** [37316210](https://pubmed.ncbi.nlm.nih.gov/37316210/). DeMatteo’s cohort reported **64%** with school problems during recovery and **31%** with grade drops even as activity stages advanced [31396150](https://pubmed.ncbi.nlm.nih.gov/31396150/)—the vignette pattern is common, not exotic.

**Decision sequence:**

1. **Write both coordinates on one line:** RTL **3** (full day + accommodations) · RTP **2A** (solo light aerobic, **not** team conditioning yet)—not “stage 2” in isolation.

2. **Gate sport advance on combined load:** If **screen or homework** provokes **>2/10 for >1 hour**, **hold RTP advance** even when the walk was tolerated. Weekly volume from pillar2-003 still matters across the week—but **today’s ceiling** may be set by schoolwork, not minutes.

3. **Separate exercise from team reintegration:** Team jogs and passing drills belong to **later RTP steps**. A walk prescription from clinic is not the same as **return to practice environment**.

4. **One communication touchpoint:** AT, counselor, and parent receive the **same sentence** (see §5 template)—not three incompatible summaries.

**Parent-facing translation:** “The walk after school was fine. The lab on the laptop was too much **the same day**. We are not failing sport—we are **pausing both tracks** until a day tolerates **walk + normal homework** without a late spike. Team practice waits.”

**Clinician note:** Purcell et al. list symptom burden, rest, age/grade, and **course load** as core return-to-school factors [29500251](https://pubmed.ncbi.nlm.nih.gov/29500251/). Halstead emphasizes school environmental adjustments, not only attendance status [24163302](https://pubmed.ncbi.nlm.nih.gov/24163302/). Aerobic prescription without those variables is a **single-track note**.

::: {.callout-caution}
**Not medical advice. Educational use only. Data from published literature.**
:::

#### 5. SOP checklist — dual-track documentation {#sop-checklist-dual-track}

*Synthesizes CISG coordination rules, DeMatteo staging, Carson implementation warnings, and pillar2-002/003 intensity/volume rules for education only [37316210](https://pubmed.ncbi.nlm.nih.gov/37316210/) [31396150](https://pubmed.ncbi.nlm.nih.gov/31396150/) [35292472](https://pubmed.ncbi.nlm.nih.gov/35292472/) [32079897](https://pubmed.ncbi.nlm.nih.gov/32079897/) [33856860](https://pubmed.ncbi.nlm.nih.gov/33856860/).*

##### One-page SOP (community default)

| Role | Checklist |
|------|-----------|
| **Clinician** | □ Record **RTL step (1–4)** and **RTP step (1–6)** on every note—not “progressing well” |
| | □ State **which track is limiting** today (school / sport / both) |
| | □ Pair RTP step 2 with **intensity path** from [pillar2-002](../pillar2-002-no-bctt-prescription-paths/index.html) (% band, RPE, or symptom cap) |
| | □ Pair weekly **minutes × days** from [pillar2-003](../pillar2-003-twenty-minutes-from-where/index.html) when prescribing aerobic volume |
| | □ **Hold RTP 4–6** until full-time school **without** concussion accommodations (Amsterdam youth gate) |
| | □ Send **one** clearance paragraph to school + coach + family with **both** labels |
| **School / counselor** | □ Accommodation list maps to **RTL step**, not generic “concussion protocol” |
| | □ Ask for **RTP step** before allowing PE, club sport, or conditioning |
| | □ Flag if athlete is in **RTL 1–3** while team resumes non-contact training |
| **Athletic trainer / coach** | □ No contact drills until **written RTP ≥4** **and** clinician documents **RTL 4** (or explicit exception) |
| | □ Log **symptoms after session**; ask about **same-day school load** before clearing next sport step |
| **Parent / athlete** | □ Home log: columns for **RTL activity**, **RTP activity**, **peak symptom**, **duration >1 h?** |
| | □ Stop same-day stacking if **combined** load triggers >2/10 for >1 h |
| | □ “Cleared to run” ≠ “cleared for games” ≠ “full school day without breaks” |

##### Example one-line status (copy to letter)

> **RTL 3** (full day + breaks, tests deferred, PE excused, reduced screens) · **RTP 2A** (solo walk 15–20 min, symptom cap +2/10) · **No team conditioning / no contact** · Re-check in 7 d or sooner if **walk + homework** same day worsens symptoms >2/10 for >1 h.

```{mermaid}
flowchart LR
 subgraph school [RTL track]
 R1[Step 1-2 home/homework] --> R3[Step 3 accommodations]
 R3 --> R4[Step 4 full day no accommodations]
 end
 subgraph sport [RTP track]
 P1[Step 1-2 light aerobic] --> P3[Step 3 individual drills]
 P3 --> P4[Steps 4-6 team/contact/play]
 end
 R4 --> P4
 P1 -.->|overlap allowed early| R3
```

*Diagram:* Early overlap (dotted) is permitted; **high-risk sport (steps 4–6)** waits on **full RTL (step 4)** per Amsterdam youth guidance [37316210](https://pubmed.ncbi.nlm.nih.gov/37316210/).

Optional intensity helper: the standalone [HR calculator](/tools/hr-calculator/) from pillar2-002 encodes **% bands**—use as a **ceiling within RTP step 2**, not as permission to ignore RTL limits.

::: {.callout-important}
**Educational use only.** Staging labels summarize published consensus and chart-review literature—they do not replace local return-to-play law, school board policy, or individual medical judgment.
:::

::: {.callout-caution}
**Not medical advice. Educational use only. Data from published literature.**
:::

#### 6. Limitations

1. **Step text paraphrased from CISG appendices;** local policies and tools (REAP, SCHOOLFirst, Living Guideline) add detail not fully tabulated here [PMID: 34777657](https://pubmed.ncbi.nlm.nih.gov/34777657/) [PMID: 35292473](https://pubmed.ncbi.nlm.nih.gov/35292473/) [PMID: 26589561](https://pubmed.ncbi.nlm.nih.gov/26589561/).

2. **North American chart-review dominance;** “premature” RTL/RTP definitions vary by chart auditor and era [24925965](https://pubmed.ncbi.nlm.nih.gov/24925965/) [35292472](https://pubmed.ncbi.nlm.nih.gov/35292472/).

3. **Marginal proportions:** Carson relapse figures are not, in published form, a full set of conditional probabilities for individual counseling.

4. **Abbreviation collision:** DeMatteo **RTS = school** vs CISG **RTS = sport** persists in the literature—readers must translate, not merge [31396150](https://pubmed.ncbi.nlm.nih.gov/31396150/) [37316210](https://pubmed.ncbi.nlm.nih.gov/37316210/).

5. **No adolescent PPCS RCT** randomizing **dual-track coordination** vs sport-track-only follow-up with preregistered combined-load outcomes.

6. **Aerobic PPCS evidence still sport-track weighted:** Leddy-style subsymptom trials enroll exercise arms without parallel school-load manipulation [30715132](https://pubmed.ncbi.nlm.nih.gov/30715132/)—pillar2-002/003 prescriptions may outpace RTL documentation.

7. **Legal scope:** Educational synthesis only—not medical advice.

##### What these limitations mean in practice

The most actionable gap is not missing a seventh RTP stage—it is missing a **shared row** in the chart. Until trials report **both** tracks with **combined daily symptoms**, clinics should treat dual-track staging as a **documentation and communication** problem first, and a **prescription** problem second. That is where pillar2-002’s bands and pillar2-003’s weekly minutes actually attach: inside labeled RTP steps, gated by RTL tolerance—not floating alone on a discharge sheet.

::: {.callout-caution}
**Not medical advice. Educational use only. Data from published literature.**
:::

Berlin paired **four RTL steps** with **six RTP steps** in one consensus document [28446457](https://pubmed.ncbi.nlm.nih.gov/28446457/). Amsterdam refined the symptom rules and made the parallel-track logic explicit [37316210](https://pubmed.ncbi.nlm.nih.gov/37316210/). Canadian charts still show **both** tracks released too early, with **RTL improvements lagging RTP** [35292472](https://pubmed.ncbi.nlm.nih.gov/35292472/). The literature is not silent on coordination—it is **easy to read one column and call it recovery**.

This is the fourth piece in Foundation Stack. If pillar2-001 mapped protocol fragmentation, pillar2-002 mapped intensity without treadmills, and pillar2-003 mapped **time without superstition**, this one maps **staging without collapsing two checklists into one**.

The [RSS feed](../../index.xml) on the hub will notify readers when the next installment publishes. The next installment works through a different unresolved problem in the same literature.

---

**Series:** Foundation Stack · **pillar2-004** · **Pillar:** methods (staging / deployability)
